Generating Random Logic Programs Using Constraint Programming

### 2020-01-10 paper/paper.tex
An example for predicate independence

@@

% TODO: define 'gap'

@@

\subsection{TODO}

@@

\section{Considering Each Clause Separately}

\subsection{Heads of Clauses}

@@

\subsection{Bodies of Clauses}

@@

Auxiliary variables: $\variable{numNodes}, \variable{numTrees} \in \{ 1, \dots,
\maxNumNodes{} \}$. The former counts the number of nodes in the main tree. The
latter counts the number of trees in total.

@@

\subsubsection{Nodes}

@@

\subsubsection{Constraints}

\begin{constraint}
  \variable{treeStructure} represents \variable{numTrees} trees, i.e.,
  \[
    \variable{tree}(\variable{treeStructure}, \variable{numTrees})\footnote{This
      constraint uses dominator-based filtering by Fages and Lorca
      \cite{DBLP:conf/cp/FagesL11}.}.
  \]
\end{constraint}

@@

\begin{constraint}
  $\variable{numTrees} + \variable{numNodes} = \maxNumNodes{} + 1$.
\end{constraint}

@@

\subsection{Eliminating Variable Symmetries} \label{sec:variable_symmetry}

@@

\begin{example}
  Let $\constants{} = \emptyset$, $\variables{} = \{ X, Y, Z \}$, $\maxArity{} =
  2$, $\maxNumNodes{} = 3$, and consider the clause
  \[
    \variable{sibling}(X, Y) \gets \variable{parent}(X, Z) \land
    \variable{parent}(Y, Z).
  \]
  Then $\variable{terms} = [X, Y, \Box, \Box, X, Z, Y, Z]$ (the boxes represent
  the conjunction node), $\variable{occurrences} = [\{ 0, 4 \}, \{ 1, 6 \},
  \{ 5, 7 \}, \{ 2, 3 \}]$, and $\variable{introductions} = [0, 1, 5]$.
\end{example}

@@

@@

\begin{definition}[Adjacency matrix representation]
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$
  defined by
  \begin{align*}
    A_{i,j} = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{definition}

@@

\begin{algorithm}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$} \times
    \getDependencies{$p_2$}$ s.t. $d_1.\predicate = d_2.\predicate$}{
    \If{$d_1$.\isDetermined{} {\bf and} $d_2$.\isDetermined{}}{
      \fail{}\;
    }
    \uIf{$d_1$.\isDetermined{}}{
      $\mathbf{A}[d_2.\source][d_2.\target]$.\removeValue{$1$}\;
    }
    \ElseIf{$d_2$.\isDetermined{}}{
      $\mathbf{A}[d_1.\source][d_1.\target]$.\removeValue{$1$}\;
    }
  }
  \caption{Propagation}
\end{algorithm}

\begin{algorithm}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$} \times
  \getDependencies{$p_2$} \mid d_1.\predicate = d_2.\predicate \}$\;
  \If{$\{ (d_1, d_2) \in D \mid d_1.\isDetermined{}, d_2.\isDetermined{} \} \ne
    \emptyset$}{
    \Return{FALSE}\;
  }
  \If{$D = \emptyset$}{
    \Return{TRUE}\;
  }
  \Return{UNDEFINED}\;
  \caption{Entailment}
\end{algorithm}

\begin{algorithm}
  \SetKwData{edgeExists}{edgeExists}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{an $n \times n$ adjacency matrix $\mathbf{A}$}
  \Fn{\getDependencies{$p$}} {
    $D \gets \{ p \}$\;
    \Repeat{$D' = D$}{
      $D' \gets D$\;
      \For{$d \in D$}{
        \For{$i \gets 1$ \KwTo $n$}{
          $\edgeExists \gets \mathbf{A}[i][d.\predicate] = \{ 1 \}$\;
          \uIf{$\edgeExists$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ i \}$\;
          }
          \uElseIf{$\edgeExists$ {\bf and not} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, d.\source, d.\target) \}$\;
          }
          \ElseIf{$|\mathbf{A}[i][d.\predicate]| > 1$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, i, d.\predicate) \}$\;
          }
        }
      }
    }
    \Return{$D$}\;
  }
  \caption{Computing the dependencies of a predicate}
\end{algorithm}

@@

@@

The author would like to thank Vaishak Belle for his comments. This work was
supported by the EPSRC Centre for Doctoral Training in Robotics and Autonomous
Systems, funded by the UK Engineering and Physical Sciences Research Council
(grant EP/S023208/1).

>>>>

\usetikzlibrary{arrows.meta}

@@

TODO: define all the relevant terminology from logic. If a predicate has arity
$n$ and it is as of yet undecided what $n$ terms will fill those spots, we will
say that the predicate has $n$ \emph{gaps}.

@@

@@

\section{Heads of Clauses}

@@

\section{Bodies of Clauses}

@@

@@

\subsection{Nodes}

@@

\subsection{Constraints}

\begin{definition}
  We define $\variable{numTrees} \in \{ 1, \dots, \maxNumNodes{} \}$ to count
  the number of trees in our representation of a clause using the
  $\variable{tree}(\variable{treeStructure}, \variable{numTrees})$\footnote{This
    constraint uses dominator-based filtering by Fages and Lorca
    \cite{DBLP:conf/cp/FagesL11}.} constraint.
\end{definition}

\begin{definition}
  For convenience, we also define $\variable{numNodes} \in \{ 1, \dots,
  \maxNumNodes{} \}$ to count the number of nodes in the main tree. We define it
  as
  \[
    \variable{numNodes} = \maxNumNodes{} - \variable{numTrees} + 1.
  \]
\end{definition}

@@

@@

\section{Eliminating Variable Symmetries} \label{sec:variable_symmetry}

@@

\begin{example}
  Let $\constants{} = \emptyset$, $\variables{} = \{ X, Y, Z \}$, $\maxArity{} =
  2$, $\maxNumNodes{} = 3$, and consider the clause
  \[
    \mathsf{sibling}(X, Y) \gets \mathsf{parent}(X, Z) \land
    \mathsf{parent}(Y, Z).
  \]
  Then $\variable{terms} = [X, Y, \Box, \Box, X, Z, Y, Z]$ (the boxes represent
  the conjunction node), $\variable{occurrences} = [\{ 0, 4 \}, \{ 1, 6 \},
  \{ 5, 7 \}, \{ 2, 3 \}]$, and $\variable{introductions} = [0, 1, 5]$.
\end{example}

@@

In this section, we define a notion of predicate independence as a way to
constrain the probability distributions defined by the generated programs. We
also describe efficient algorithms for propagation and entailment checking.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw] (father) at (0, 0.5) {$\mathsf{father}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (parent) at (2, 0) {$\mathsf{parent}$};
    \node[draw] (sibling) at (4, 0) {$\mathsf{sibling}$};
    \draw[-{Stealth[scale=1.5]}] (father) -- (parent);
    \draw[-{Stealth[scale=1.5]}] (mother) -- (parent);
    \draw[-{Stealth[scale=1.5]}] (parent) -- (sibling);
  \end{tikzpicture}
  \caption{The predicate dependency graph of the program in \cref{ex:program}}
  \label{fig:predicate_dependencies}
\end{figure}

\begin{example} \label{ex:program}
  Consider the following program:
  \begin{align*}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z),\\
    \mathsf{parent}(X, Y) &\gets \mathsf{father}(X, Y) \lor \mathsf{mother}(X, Y).
  \end{align*}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}. We can
  then list the dependencies of each predicate:
  \begin{alignat*}{3}
    D_{\mathsf{father}} &= \{ \mathsf{father} \}, \quad && D_{\mathsf{parent}} &&= \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent} \}, \\
    D_{\mathsf{mother}} &= \{ \mathsf{mother} \}, \quad && D_{\mathsf{sibling}} &&= \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent}, \mathsf{sibling} \}.
  \end{alignat*}
  Hence, the only pair of independent predicates in this program is
  $\mathsf{father}$ and $\mathsf{mother}$.
\end{example}

% TODO: continue writing from here
\begin{definition}[Adjacency matrix representation]
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$
  defined by
  \begin{align*}
    A_{i,j} = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{definition}

@@

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$} \times
    \getDependencies{$p_2$}$ s.t. $d_1.\predicate = d_2.\predicate$}{
    \If{$d_1$.\isDetermined{} {\bf and} $d_2$.\isDetermined{}}{
      \fail{}\;
    }
    \uIf{$d_1$.\isDetermined{}}{
      $\mathbf{A}[d_2.\source][d_2.\target]$.\removeValue{$1$}\;
    }
    \ElseIf{$d_2$.\isDetermined{}}{
      $\mathbf{A}[d_1.\source][d_1.\target]$.\removeValue{$1$}\;
    }
  }
  \caption{Propagation}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$} \times
  \getDependencies{$p_2$} \mid d_1.\predicate = d_2.\predicate \}$\;
  \If{$\{ (d_1, d_2) \in D \mid d_1.\isDetermined{}, d_2.\isDetermined{} \} \ne
    \emptyset$}{
    \Return{FALSE}\;
  }
  \If{$D = \emptyset$}{
    \Return{TRUE}\;
  }
  \Return{UNDEFINED}\;
  \caption{Entailment}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwData{edgeExists}{edgeExists}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{an $n \times n$ adjacency matrix $\mathbf{A}$}
  \Fn{\getDependencies{$p$}} {
    $D \gets \{ p \}$\;
    \Repeat{$D' = D$}{
      $D' \gets D$\;
      \For{$d \in D$}{
        \For{$i \gets 1$ \KwTo $n$}{
          $\edgeExists \gets \mathbf{A}[i][d.\predicate] = \{ 1 \}$\;
          \uIf{$\edgeExists$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ i \}$\;
          }
          \uElseIf{$\edgeExists$ {\bf and not} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, d.\source, d.\target) \}$\;
          }
          \ElseIf{$|\mathbf{A}[i][d.\predicate]| > 1$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, i, d.\predicate) \}$\;
          }
        }
      }
    }
    \Return{$D$}\;
  }
  \caption{Computing the dependencies of a predicate}
\end{algorithm}

@@

\section{Conclusion \& Future Work}

@@

% The author would like to thank Vaishak Belle for his comments.
This work was supported by the EPSRC Centre for Doctoral Training in Robotics
and Autonomous Systems, funded by the UK Engineering and Physical Sciences
Research Council (grant EP/S023208/1).
